---
title: "Section 1.3: Interactive multivariate visualization with shiny"
author: "Haeun Lee"
date: "2026-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Interactive multivariate visualization with shiny

## Graphical user interfaces with Bioconductor and shiny
Introducing the shiny-based GUI concept for multivariate analysis

```{r setup,echo=FALSE,results="hide"}
suppressWarnings({
suppressMessages({
suppressPackageStartupMessages({
library(shiny)
library(cluster)
library(ph525x)
})
})
})
```

### Example

```{r lkb,eval=FALSE}
mtcars    # built-in in R
pairs(mtcars[,1:5])    # view relationships between first 5 variables 

# open an interactive pair plot and clustering visualization
library(ph525x)
dfHclust(mtcars)

# repeat with a different built-in data frame
dfHclust(iris)
```

## Programming shiny

In this subsection we will break up the main `dfHclust` function and explain its elements.  The code was written in a very naive manner but even so has three virtues:

* it is relatively short and self-contained
* it works and does something useful in interactive data exploration
* it is easy to extend by replicating and modifying short subparts

Some aspects that are likely to need improvement

* handling of global variables for use in the `ui` component
* excessive repetition of data.frame subsetting in the `server` component

### Starting out

We have a simple fixed interface and fail if we don't get a data.frame with at least two columns.  We fail if the necessary software is not in place.
```
dfHclust = function(df) {
# validate input
   stopifnot(inherits(df, "data.frame"))
   stopifnot(ncol(df)>1)
# obtain software
   require(shiny)
   require(cluster)
```
### Some vectors to populate interface option sets

```
# global variables ... 
   nms = names(df)
   cmeths = c("ward.D", "ward.D2",
             "single", "complete", "average", "mcquitty",
             "median", "centroid")
   dmeths = c("euclidean", "maximum", "manhattan", "canberra",
             "binary")
```

### The layout of the user interface

There are several high-level options that can control
the browser interface.  We'll use a flexible approach
called `fluidPage`; see also the shinydashboard package.

```
#
# main shiny components: ui and server
#   ui: defines page layout and components
#   server: defines operations
#
   ui <- fluidPage(
#
# we will have four components on sidebar: selectors for 
# distance, agglomeration method, height for tree cut, and variables to use
```

#### Title and sidebar
```
#
     titlePanel(paste(substitute(df), "hclust")),
     sidebarPanel(
          helpText(paste("Select distance:" )),
          fluidRow(
             selectInput("dmeth", NULL, choices=dmeths,
               selected=dmeths[1])),
          helpText(paste("Select clustering method:" )),
          fluidRow(
             selectInput("meth", NULL, choices=cmeths,
               selected=cmeths[1])),
          helpText(paste("Select height for cut:" )),
          fluidRow(
             numericInput("cutval", NULL, value=40, min=0, max=Inf, step=1)),
          helpText(paste("Select variables for clustering from", substitute(df), ":" )),
          fluidRow(
             checkboxGroupInput("vars", NULL, choices=nms,
               selected=nms[1:2]))
            ),
```
#### Tabs for the primary displays

```
#
# main panel is a simple plot
#
     mainPanel(
       tabsetPanel(
        tabPanel("tree", 
         plotOutput("plot1")),
        tabPanel("pairs", 
         plotOutput("pairsplot")),
        tabPanel("silh", 
         plotOutput("silplot"))
         )
       )
  )  # end fluidPage
```

### Defining the server
   
```
#
# a function with up to three arguments (input, output, session)
# can be used to define the server component of the app
# renderPlot makes it reactive, so when input components are altered,
# data frame in use and plot are updated
#
   server <- function(input, output) {
     output$plot1 <- renderPlot({
       xv = df[,input$vars]
       plot(hclust(dist(data.matrix(xv),method=input$dmeth), method=input$meth),
         xlab=paste(input$dmeth, "distance;", input$meth, "clustering"))
       abline(h=input$cutval, lty=2, col="gray")
     })
     output$pairsplot <- renderPlot({
       xv = df[,input$vars]
       pairs(data.matrix(xv))
     })
     output$silplot <- renderPlot({
       xv = df[,input$vars]
       dm = dist(data.matrix(xv),method=input$dmeth)
       hc = hclust(dist(data.matrix(xv),method=input$dmeth), method=input$meth)
       ct = cutree(hc, h=input$cutval)
       plot(silhouette(ct, dm))
     })
   }
   
   shinyApp(ui, server)
}
```

### Some gory details

A "shiny app" consists of two main components, a user interface and
a server function.  To use the infrastructure the `shiny`
library must be attached to an R session.
The application can be started in various ways; we use
`shinyApp` and supply two arguments, `ui` and `server`.

* `ui` is an instance of `shiny.tag.list`.  You can get a feel
for this by inspecting the result of `fluidPage()`.

* `server` is a function of three arguments `input`, `output`
and `session`; the latter is optional.  `input` is a list
with bindings given values in the `ui` component, and `output`
will be populated with elements in the server for rendering in
the UI.

## Experimentally modifying a shiny app in Rstudio

```{r lkb,eval=FALSE}
library(ph525x)
dfh2 = dfHclust    # copy of dfHclust

fix(dfh2)
# apply following edit in the interactive window
# helpText(paste("Select variables for clustering from", 
#                substitute(df), ":")), fluidRow(selectInput("vars",     # replace checkboxInput with selectInput
#                NULL, choices = nms, selected = nms[1:2], multiple=TRUE)))    # add multiple argument


dfHclust(mtcars)    # checkboxes for variables
dfh2(mtcars)    # checkboxes replaced with dropdown/search
```

## Multivariate visualization with gene expression data

See "Section1.3_esHclust.Rmd"

## Closing remarks on genome-scale visualization

See "Section1.3_finalViz.Rmd"

## Shiny for gene expression clustering: Assessment

### Setup of the expression data and app

```{r}
library(Biobase)
library(hgu133a.db)
library(ph525x)

esHclust = function(es) {
  emat = t(exprs(es))
  rownames(emat) = sampleNames(es)
  dd = data.frame(emat)
  dfHclust(dd)
}

library(tissuesGeneExpression)
data(tissuesGeneExpression)
tgeES = ExpressionSet(e)
annotation(tgeES) = "hgu133a.db"
pData(tgeES) = tab
featureNames(tgeES) =
  make.names(mapIds(hgu133a.db, keys=featureNames(tgeES),
    keytype="PROBEID", column="SYMBOL"), unique=TRUE)
sampleNames(tgeES) = make.names(tgeES$Tissue, unique=TRUE)
```

How many of the feature names of tgeES are missing (NA)?

```{r}
sum(featureNames(tgeES) == "NA")
sum(is.na(rownames(exprs(tgeES))))
```

The originating data are limited to probes with gene identifiers. The make.names call ensures that duplicates get unique suffixes of the form .[n].

### Verifying initial configuration

Run esHclust(tgeES[1:50,1:50]). Set the cutoff height to 2.

What is the average silhouette value for this clustering/partition? 0.43

```{r}
esHclust(tgeES[1:50,1:50])

# OR

ex = exprs(tgeES[1:50,1:50])
h = hclust(dist(t(ex[1:2,])), method="ward.D")
cc = cutree(h, h=2) 
summary(silhouette(cc, dist(t(ex[1:2,]))))
```

### The 50-gene signature for tissue differentiation

Instead of using the first 50 genes, exploit the ExpressionSet interface to make a targeted investigation. We used limma to identify 50 genes for which a null hypothesis of common mean expression across tissues is rejected. The gene set is identified via

```{r}
library(limma)
mm = model.matrix(~Tissue, data=pData(tgeES))
f1 = lmFit(tgeES, mm)
ef1 = eBayes(f1)
sig50 = rownames(ef1$coef[order(ef1$F,decreasing=TRUE)[1:50],])
```

Now we can run

```{r}
esHclust( tgeES[sig50,] )
```

Set the height for cut to 2. The average silhouette for the top 2 genes, Euclidean distance, and ward.D clustering is 0.35.

How many clusters are formed by cutting the tree at 2?

7

## Genes ranked 49th and 50th

Scroll to the bottom of the feature list and check the bottom two genes: EEF1G.2 and KLRC4. Then uncheck the top two genes RPL37A and RPL41. Change the clustering method to ward.D2

What is the average silhouette value for the clustering based on these choices? 0.55

```{r}
ex = exprs(tgeES[sig50,])
h = hclust(dist(t(ex[49:50,])), method="ward.D2")
cc = cutree(h, h=2)
summary(silhouette(cc, dist(t(ex[49:50,]))))
```

### Random forests for the 5 gene signature

In the book chapter we showed how the random forests algorithm could be used to assess the set of 50 genes with greatest F values for discriminating tissue-specific means. Try it with the 5 gene signature:

```{r}
sig5 = c("IL26", "ZNF674", "UBC.1", "C7orf25.1", "RPS13")
library(MLInterfaces)
library(randomForest)
set.seed(1234)
rf2 = MLearn(Tissue~., tgeES[sig5,], randomForestI, xvalSpec("NOTEST"))
RObject(rf2)
```

In percent, what is the out of bag estimate of the misclassification error rate? 37.04%









