---
layout: page
title: "Some final comments on genome-scale visualization"
---

```{r options, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
suppressWarnings({
library(BiocStyle)
library(Biobase)
library(randomForest)
library(MLInterfaces)
library(tissuesGeneExpression)
library(limma)
library(ph525x)
library(RCircos)
})
})
```


## RCircos

RCircos is not distributed in Bioconductor, but can
be useful for developing compact displays of interactions
among genomic elements.  I am unaware of any interfaces between
Bioconductor data classes and RCircos, and this topic deserves
attention.

In the ph525x package we have added a selection of trans-eQTL
findings from Westra et al. Nature 2013 (doi: 10.1038/ng.2756).
We show a few SNP-gene associations from this study:
```{r lksn,fig=TRUE,message=FALSE,fig.height=9,fig.width=9}
library(ph525x)
library(SNPlocs.Hsapiens.dbSNP144.GRCh37)
library(BSgenome)
data(westraTransSel)
westraTransSel[1:3]
sglToCircos(westraTransSel[1:5])
```

```{r}
library(circlize)

# Prepare data
links <- do.call(rbind, lapply(names(westraTransSel[1:5]), function(snp) {
  genes <- westraTransSel[[snp]]
  data.frame(snp = snp, gene = genes)
}))

snps <- unique(links$snp)
genes <- unique(links$gene)

all_nodes <- c(snps, genes)

# Initialize circle with spacing
circos.clear()
circos.par(gap.degree = 4)

circos.initialize(factors = all_nodes,
                  xlim = cbind(rep(0, length(all_nodes)), rep(1, length(all_nodes))))

# ADD TRACK
circos.trackPlotRegion(ylim = c(0, 1), panel.fun = function(x, y) {
  sector.name <- get.cell.meta.data("sector.index")
  
  circos.text(
    x = 0.5, y = 0.5,
    labels = sector.name,
    cex = 0.6,
    facing = "clockwise",
    niceFacing = TRUE
  )
}, bg.border = NA)

# Draw links
for (i in 1:nrow(links)) {
  circos.link(
    links$snp[i], 0.5,
    links$gene[i], 0.5,
    col = "#4682B480"
  )
}

grid.col <- c(
  setNames(rep("red", length(snps)), snps),
  setNames(rep("blue", length(genes)), genes)
)

circos.trackPlotRegion(
  ylim = c(0, 1),
  bg.col = grid.col,
  panel.fun = function(x, y) {
    circos.text(get.cell.meta.data("xcenter"), 0.5,
                get.cell.meta.data("sector.index"),
                cex = 0.5, facing = "clockwise", niceFacing = TRUE)
  }
)
```


## ComplexHeatmap

`r Biocpkg("ComplexHeatmap")` has a very nice vignette addressing many issues in combining heatmaps and repurposing the heatmap concept.  The oncoprint example in the vignette is particularly comrelling.  To use this interactively with TCGA, contact [the ISB](http://isb-cancer-genomics-cloud.readthedocs.io/en/latest/sections/FAQ.html) and obtain a cloud platform account.
Then obtain the `r Biocpkg("cgcR")` package, load it, and run `isbApp()`.
You will have to authenticate with google to get access to the BigQuery representation of TCGA.

```{r}
# BiocManager::install("ComplexHeatmap")
# BiocManager::install("TCGAbiolinks")
library(ComplexHeatmap)


library(TCGAbiolinks)

query <- GDCquery(
  project = "TCGA-BRCA",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification"
)

GDCdownload(query)
data <- GDCprepare(query)

# Old version
# library(cgcR)
# isbApp()
```

## WebGL and interaction with data

In the short concluding video we use the MLInterfaces plspinHcube function to illustrate several aspects of interactivity: GUI for tuning, mouse-controlled rotation, and mouseover for point interrogation.

```{r}
library(MLInterfaces)
library(threejs)
library(mlbench)
plspinHcube()
```

## EpiViz

The `r Biocpkg("epivizr")` package interacts with the [epiviz](https://epiviz.github.io/) system and is capable of substantial feats of data integration and higher-level data interactivity.
