---
title: "Section 5: Inference for genomics with Bioconductor"
author: "Haeun Lee"
date: "2026-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
BiocManager::install(c("SpikeInSubset",
                       "genefilter",
                       "qvalue",
                       "limma",
                       "genomicsclass/maPooling",
                       "leukemiasEset",
                       "preprocessCore"))
```

# Differential expression and multiple comparisons
## Statistical prerequisites

## Biological versus technical variability
The t-test takes into account within population variability. In experimental biology, this variability can be divided into two main classes: technical and biological.

## t-tests in genomics
```{r}
library(genefilter)
# rowttests(assay_data, grouping_factor)
```

## Moderated t-tests with the limma package
Three steps in the basic limma analysis:
- lmFit() to fit linear models
- eBayes() to perform hierarchical modeling
- topTable() to return the top features

### Assessment
```{r}
BiocManager::install("genomicsclass/maPooling")

library(Biobase)
library(maPooling)
data(maPooling)
pd=pData(maPooling)
pooled=which(rowSums(pd)==12)
individuals=which(rowSums(pd)==1)
##remove replicates
individuals=individuals[-grep("tr",names(individuals))]
pool = exprs(maPooling)[,pooled] # from maPooling@assayData
indiv = exprs(maPooling)[,individuals] # from maPooling@assayData
strain = ifelse(grepl("a",rownames(pData(maPooling))),0,1)

g_pool = strain[pooled]
g_indiv = strain[individuals]

technicalsd <- rowSds(pool[,g_pool==1])
biologicalsd <- rowSds(indiv[,g_indiv==1])

mean(biologicalsd>technicalsd)

## plot
plot(technicalsd,biologicalsd)
abline(0,1,col=2)

tt = rowttests(pool,factor(g_pool))
library(qvalue)
qvals = qvalue(tt$p.value)$qvalue
sum(qvals < 0.05)
i = which(qvals < 0.05)

# Much larger than we expected given that we estimated an FDR of 5% in the first list. This is because the pooled data is not accounting for biological variability.
pvals_ind = rowttests(indiv,factor(g_indiv))$p.value
mean(pvals_ind[i]>0.05)

qvals_ind = qvalue(pvals_ind)$qvalue
i_ind = which(qvals_ind < 0.05)

library(limma)
fit <- lmFit(indiv, design=model.matrix(~ g_indiv))
fit <- eBayes(fit)

qval_modt = qvalue(fit$p.value[,2])$qvalue
## qval_modt = qvalue(fit$p.value)$qvalue[,2] --> this give another value!
# all p-values from all columns are used >> one global π₀ is estimated from the full set

mean(qval_modt[i_ind]<0.05)
# Among the indices in i_ind, what fraction are significant under qval_modt?
# depends on the size of i_ind

## i_ind_modt = which(qval_modt < 0.05)
## mean(i_ind_modt %in% i_ind)
# Among the discoveries from qval_modt, what fraction are also in i_ind?
# depends on the size of i_ind_modt

# Note that we get very strong agreement because N is 12 here and thus the standard error estimates are not shrunk too much.
```


# Gene set analysis: inference on coordinated expression changes
Testing for differences in predefined sets of genes between sample groups

## Gene sets
## Summary statistics for gene sets

The t-statistics from individual genes within a gene set can have positive, 0, or negative correlation.

The t-statistics can have any kind of positive, 0, or negative correlation, depending on the nature of the experiment and the genes in the set. Positive correlations can occur for example in the case of batch effects. If the genes in a set are similarly affected by batch, and batch effects are not corrected for, then a high t-statistic for one gene means that a high t-statistic for the other gene is more likely. Positive correlation can also occur, for example if one gene positively regulates the other gene (in terms of gene regulatory networks). Genes within a set can also potentially have a negative correlation if one gene in a set is a negative regulator of another gene in the set. This variation could then lead to a negative correlation between the expression of these two genes, and therefore a negative correlation between the two t-statistics.

## Hypothesis testing for gene sets
In real workflows:
	•	limma → t-statistics → CAMERA
	•	DESeq2 → Wald stats → GSEA
	•	edgeR → likelihood ratio stats → ROAST

## Gene set testing in R part I: Applying the roast algorithm for a single gene set
## Gene set testing in R part II

# Verified assessment

## Q3: Exploring dependence
```{r}
var(rowMeans(matrix(rnorm(10000 * 10, 0, 1), ncol=10)))

library(MASS)
# 10 Normal variables with mean 0 and variance 1, but each random variable has a correlation of 0.7 with the other variables. We can create such random variables in R using the multivariate normal distribution:
Sigma = matrix(.7, ncol=10, nrow=10)
diag(Sigma) = 1
mvrnorm(n=1,mu=rep(0,10),Sigma=Sigma)
```

$$
\begin{eqnarray}
Var(\bar{t}) &=& \frac{1}{N^2} \mbox{var}( (1 \dots 1) (t_1 \dots
t_N)' ) \\
&=& \frac{1}{N^2}(1 \dots 1)
\begin{pmatrix}
1 & \rho & \dots & \rho & \rho \\
\rho & 1 & \rho & \dots  & \rho \\
\dots & \dots & \dots & \dots & \dots \\
\rho &  \rho & \dots & \rho & 1 \\
\end{pmatrix}
(1 \dots 1) ' \\
&=& \frac{1}{N^2}\{N + (N-1) N \rho \} \\
&=& \frac{1}{N}\{1 + (N-1) \rho \} \\
\end{eqnarray}
$$
```{r}
avgcorrs = 1/10 * (1 + (10 - 1) * 0.7) # variance of the average of these 10 random variables

##Using simulation:
set.seed(2)
var(rowMeans(mvrnorm(n=10000,mu=rep(0,10),Sigma=Sigma)))

```

## Q4: Gene set analysis with the roast algorithm
```{r}
library(GEOquery)
e = getGEO("GSE34313")[[1]]

e$condition = as.factor(e$characteristics_ch1.2)
levels(e$condition) = c("dex24","dex4","control")
table(e$condition)
## Note that with ExpressionSets we can access the phenoData columns directly
## Specifically, note that e$condition is equivalent to pData(e)$condition

names(fData(e))
fData(e)$GO_ID[1:4] # data includes GO terms for every gene

# compare the control samples to those treated with dexamethasone (the hormone) after 4 hours
lvls = c("control", "dex4")
es = e[,e$condition %in% lvls]
es$condition = factor(es$condition, levels=lvls)

library(limma)
library(qvalue)
design = model.matrix(~ es$condition)
fit = lmFit(es, design=design)
fit = eBayes(fit)
topTable(fit)[,c(6,7,18,22)]
```
Top genes are common immune-response genes (CSF2, LIF, CCL2, IL6). Also present is FKBP5, a gene which regulates and is regulated by the protein which receives the glucocorticoid hormone

ROAST method for gene set testing
We can test a single gene set by looking up the genes which contain a certain GO ID, and providing this to the roast() function.

The roast function performs an advanced statistical technique, rotation of residuals, in order to generate a sense of the null distribution for the test statistic. The test statistic in this case is the summary of the scores from each gene. The tests are self-contained because only the summary for a single set is used, whereas other gene set tests might compare a set to all the other genes in the dataset, e.g., a competitive gene set test.

The result here tells us that the *immune response* genes are significantly down-regulated, and additionally, mixed up and down.
```{r}
# Immune response
set.seed(1)
idx = grep("GO:0006955", fData(es)$GO_ID)
length(idx)
r1 = roast(es, idx, design)
r1

set.seed(2)
idx2 = grep("GO:0045454", fData(es)$GO_ID)
length(idx2)
r2 = roast(es, idx2, design)
r2
r2[1]$p.value$P.Value[1]
```

## Q5: Inference for multiple gene sets

We can also use the mroast() function to perform multiple roast tests. First we need to create a list, which contains the indices of genes in the ExpressionSet for each of a number of gene sets. We will use the org.Hs.eg.db package to gather the gene set information. The following code used R to extracts and organizes GO annotations from a somewhat messy annotation provided by the manufacturer.
```{r}

library(org.Hs.eg.db)
org.Hs.egGO2EG
go2eg = as.list(org.Hs.egGO2EG)
head(go2eg)
```

The following code unlists the list, then gets matches for each Entrez gene ID to the index in the ExpressionSet. Finally, we rebuild the list.
```{r}
govector = unlist(go2eg)
golengths = sapply(go2eg, length)
head(fData(es)$GENE) # Entrez gene identifiers
  
idxvector = match(govector, fData(es)$GENE);table(is.na(idxvector))
##This is the organized list of indexes for genes per GO term:
idx = split(idxvector, rep(names(go2eg), golengths))
##We can see the genes like this:
go2eg[[1]]
fData(es)$GENE[idx[[1]]]
```

We need to clean this list such that there are no NA values. We also clean it to remove gene sets which have less than 10 genes.
```{r}
idxclean = lapply(idx, function(x) x[!is.na(x)])
idxlengths = sapply(idxclean, length)
idxsub = idxclean[idxlengths > 10]
length(idxsub)
```

With these indexes in place we are now ready to run mroast. The following line of code runs the multiple ROAST test. This can take about 3 minutes.
```{r}
set.seed(1)
r2 = mroast(es, idxsub, design)
head(r2)
```

Find the GO term that has the the largest proportion of upregulated genes. How many genes does this term have?
```{r}
r2[order(r2$PropUp, decreasing = TRUE), ]
r2[which.max(r2$PropUp),1]

# Note that this is a relatively small gene set. In fact, we can see that there is much more variability in this proportion for smaller gene sets:
plot(log2(r2$NGenes),r2$PropUp)
plot(log2(r2$NGenes),-log10(r2$PValue))
```
## Q6: Filtering gene sets
```{r}
idxsub2 = idxclean[idxlengths >= 50]
length(idxsub2)

set.seed(2)
r3 = mroast(es, idxsub2, design)
head(r3)

r3[order(r3$PropUp, decreasing = TRUE), ]

BiocManager::install("GO.db")
library(GO.db)
library(AnnotationDbi)
columns(GO.db)
keytypes(GO.db)
r3tab <- AnnotationDbi::select(GO.db, keys="GO:0017147",
                columns= "TERM", 
                keytype="GOID")
```
# Verified assessment: Inference for genomics
```{r}
library(leukemiasEset)
library(preprocessCore)
library(genefilter)
library(qvalue)
library(org.Hs.eg.db)

data(leukemiasEset)
```

This dataset contains gene expression data from bone marrow of 60 patients with 1 of 4 different leukemia types (ALL, AML, CLL, CML) or controls with no leukemia (NoL):
```{r}
table(leukemiasEset$LeukemiaType)
```


Before analysis, use quantile normalization on the expression data:
```{r}
boxplot(exprs(leukemiasEset))    # note different distributions across samples
lenorm = normalize.quantiles(exprs(leukemiasEset))    # quantile normalization
rownames(lenorm) = rownames(exprs(leukemiasEset))    # add row names
colnames(lenorm) = colnames(exprs(leukemiasEset))    # add column names
exprs(leukemiasEset) = lenorm    # overwrite initial exprs
boxplot(exprs(leukemiasEset))    # all samples are quantile normalized
```
Subset the leukemiasEset data to include only samples from chronic myeloid leukemia, CML ("CML"), and normal samples ("NoL") into a new ExpressionSet, cml_normal:
```{r}
cml_normal = leukemiasEset[, leukemiasEset$LeukemiaType %in% c("CML", "NoL")]
cml_normal$LeukemiaType = factor(cml_normal$LeukemiaType)
```

The reference level for a factor is always the first listed level of the factor. By default, factors are ordered alphabetically. We want to find fold changes in genes between normal samples and CML samples, so we set the reference level to be our normal samples:
```{r}
cml_normal$LeukemiaType = relevel(cml_normal$LeukemiaType, "NoL")
factor(cml_normal$LeukemiaType)    # levels are now NoL, CML
```

Note that the feature data is currently empty except for row names, which are ENSEMBL IDs. This code maps the ENSEMBL IDs to gene symbols and ENTREZ IDs and adds them to the feature data for later analysis:
```{r}
library(org.Hs.eg.db)
symbols = mapIds(org.Hs.eg.db, rownames(cml_normal), keytype = "ENSEMBL", column="SYMBOL")
ids = mapIds(org.Hs.eg.db, rownames(cml_normal), keytype = "ENSEMBL", column="ENTREZID")
fData(cml_normal)$symbol = symbols
fData(cml_normal)$gene_id = ids
fData(cml_normal)
```

## Q1: Simple t-tests for differentially expressed genes
```{r}
library(genefilter)
rtt=rowttests(exprs(cml_normal),cml_normal$LeukemiaType)
sum(rtt$p.value < 0.05)

library(qvalue)
qv = qvalue(rtt$p.value)$qvalue
sum(qv < 0.05)

```

## Q2: Relationship between p-values and q-values

Recall that the q-value gives the proportion of observations at or below that q-value that are expected to be false positives.

Make a data frame of p-values and their matching q-values, order by increasing p-value, then filter the data frame to contain only genes with p-values below .05:
```{r}
df = data.frame(p = rtt$p.value, q = qv)
df = df[rev(order(df$p)),]     # order by decreasing p-value
df_pt05 = df[df$p < .05,]    # keep only p-values below .05

df_pt05$q[1] # q-value corresponds to a p-value of .05 in this analysis
sum(rtt$p.value < .05) * df_pt05$q[1] # the number of genes in this dataset with a p-value below .05 exepected to be false positives
```

## Q3: Moderated t-tests with limma

Use the limma package to find which genes are differentially expressed between CML and normal bone marrow.

Inspect the p-values from the empirical Bayes moderated t-tests. Then, use the qvalue library to calculate the q-values for the p-values from the empirical Bayes model.
```{r}
library(limma)
fit <- lmFit(cml_normal, design=model.matrix(~ cml_normal$LeukemiaType))
colnames(coef(fit))
fit <- eBayes(fit)

pvals = fit$p.value[,2]
sum(pvals < 1e-5)

qval = qvalue(pvals)$qvalue
sum(qval < 1e-5)
```

## Q4: Top differentially expressed gene from moderated t-tests
```{r}
tt <- topTable(fit, coef=2)
tt

tt[which.min(tt$adj.P.Val),] # gene with lowest adjusted p-value


```
## Q5: Gene set analysis: mapping ENTREZ IDs to GO pathways

Use org.Hs.egGO2EG from the org.Hs.eg.db library to map GO pathways to the ENTREZ IDs of genes in those pathways:
```{r}
# convert go2eg to vector and record lengths of gene sets
go2eg = as.list(org.Hs.egGO2EG)
govector = unlist(go2eg)
golengths = sapply(go2eg, length)

# match pathway gene IDs and split into a new list
idxvector = match(govector, fData(cml_normal)$gene_id)
idx = split(idxvector, rep(names(go2eg), golengths))
head(idx)    # list of GO pathways with corresponding gene IDs

# remove NAs and keep only gene sets with over 10 genes
idxclean = lapply(idx, function(x) x[!is.na(x)])
idxlengths = sapply(idxclean, length)
idxsub = idxclean[idxlengths > 10]
```

Subset the cml_normal ExpressionSet to a single GO process, for example GO:0000002, with this code:
```{r}
process = idxsub$`GO:0000002`
idx = fData(cml_normal)$gene_id %in% process
cml_normal[idx,]

fData(cml_normal[idx,])$symbol
```

## Q6: Gene set enrichment analysis for a single gene set

Set the seed to 1, then use roast() to perform gene set enrichment analysis for the gene set GO:0034765.
```{r}
set.seed(1)
design <- model.matrix(~ cml_normal$LeukemiaType)

process = idxsub$`GO:0034765`
idx2 = fData(cml_normal)$gene_id %in% process
cml_normal[idx2,]

r1 <- roast(cml_normal, idx2, design)
r1
```
## Q7: Gene set enrichment analysis for multiple gene sets

Set the seed to 1 again, then use mroast() to perform gene set enrichment analysis for all gene sets.
```{r}
# set.seed(1)
# r2 <- mroast(cml_normal, idxsub, design)
# r2
# 
# r2_up = r2[r2$Direction == "Up",]
# sum(r2_up$PValue < 0.05)
# nrow(r2[r2$Direction == "Up" & r2$PValue < .05,])
# 
# r2_down = r2[r2$Direction == "Down",]
# sum(r2_down$PValue < 0.05)
# nrow(mr[mr$Direction == "Down" & mr$PValue < .05,])

set.seed(1)
r2 <- mroast(cml_normal, idxsub, design)

r2_up <- r2[r2$Direction == "Up", ]
sum(r2_up$PValue < 0.05)

r2_down <- r2[r2$Direction == "Down", ]
sum(r2_down$PValue < 0.05)
nrow(r2_down[r2_down$PValue < 0.05, ])
```





