---
title: "Section 2: Bioconductor basics with GRanges and Biostrings"
author: "Haeun Lee"
output: html_document
date: "2025-12-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ERBS)
library(GenomicRanges)
library(S4Vectors)
data(GM12878)
data(HepG2)
GM12878 <- updateObject(GM12878)
HepG2   <- updateObject(HepG2)
```

# IRanges and GRanges
```{r}
# Bioconductor and GitHub packages
BiocManager::install(c("Homo.sapiens",
                       "GenomicFeatures",
                       "genomicsclass/ERBS",
                       "genomicsclass/ph525x"))
```

## Motivation and introduction
```{r}
library(devtools)
# install ERBS
install_github("genomicsclass/ERBS")
library(ERBS)

BiocManager::install("updateObject")
library(updateObject)

# load GM12878 and HepG2 objects from ERBS package
library(ERBS)
data(GM12878)
data(HepG2)

# Upgrade them to the current GRanges class
GM12878 <- updateObject(GM12878)
HepG2   <- updateObject(HepG2)

# Check that they are now valid GRanges
validObject(GM12878)
validObject(HepG2)

data(HepG2)
HepG2
```

```{r}
# inspect HepG2 GRanges object
class(HepG2) # class = GRanges, package = GenomicRanges
HepG2
values(HepG2)
```

## Introduction to using GenomicRanges
```{r}
# load GM12878 and HepG2 objects from ERBS package
library(ERBS)
data(GM12878)
data(HepG2)

# seqnames extracts chromosome names
seqnames(HepG2)     # stored as type Rle
chr = seqnames(HepG2)
as.character(chr)    # view as character type

# make a table of numbers of sequences on each chromosome
table(chr)
table(chr)[1:24]    # restrict to autosomes, X and Y

# GRanges can be subsetted and ordered
HepG2[chr=="chr20",]
x = HepG2[order(HepG2),]
seqnames(x)     # demonstrate usefulness of Rle type
as.character(seqnames(x))
```

### Assessment
```{r}
median(mcols(HepG2)$signalValue)
i = which.max(mcols(HepG2)$signalValue)
seqnames(HepG2[i])

HepG2[chr=="chr16",]
# OR
chr = seqnames(HepG2)
table(chr)[16]

hist(width(HepG2),nclass=25)
median(width(ranges(HepG2)))
```

## Interval ranges: IRanges
```{r}
ir <- IRanges(101, 200)
ir * 2 # half the width

narrow(ir, start=20)
ir + 25

ir1 <- IRanges(c(1,11,21), c(3,15,27))
sum(width(ir1))

install_github("genomicsclass/ph525x")
library(ph525x)
x <-IRanges(c(101,106,201,211,221,301,306,311,351,361,401,411,501),c(150,160,210,270,225,310,310,330,390,380,415,470,510))
plotRanges(x)
gaps(x)
sum(width(gaps(x)))
disjoin(x)

par(mfrow=c(2,1))
plotRanges(x, xlim=c(0,600))
resize(x,1) # resizes the ranges to the specified width where either the start, end, or center is used as an anchor
```

### Assessment
```{r}
x = GRanges("chr1", IRanges(c(1,101),c(50,150)), strand=c("+","-"))
ranges(x)

plotGRanges = function(x) plotRanges(ranges(x))

par(mfrow=c(2,1))
plotGRanges(x)
plotGRanges(resize(x,1))
```

## Interval ranges: IRanges
### Assessment
```{r}
x = GRanges("chr1", IRanges(c(101,201,401,501),c(150,250,450,550)), strand="+")
y = GRanges("chr1", IRanges(c(101,221,301,401,541),c(150,250,350,470,550)), strand="+")

par(mfrow=c(3,1))
plotRanges(x)
plotRanges(y)
plotRanges(disjoin(c(x,y)))

GRangesList(x,y)
c(x,y)

disjoined = disjoin(c(x,y))
in.both = disjoined %over% x & disjoined %over% y
overlap.width = sum(width(disjoined[ in.both ]))

# par(mfrow=c(2,1))
# plotRanges(c(x,y)[x %over% y])
# plotRanges(disjoin(c(x,y)[x %over% y]))
# sum(width(disjoin(c(x,y)[x %over% y])))
# width(disjoin(c(x,y)[x %over% y]))

total.width = sum(width(c(x,y)))
total.width - overlap.width * 2

disjoined = disjoin(c(x,y))
not.in.both = !(disjoined %over% x & disjoined %over% y)
sum(width(disjoined[ not.in.both ]))
```

### Additional
```{r}
z = range(ranges(x))
z.GRanges = GRanges("chr1", z, strand="-")
sum(x %over% z.GRanges) # %over% is specific to 'strand'

plotRanges(x)
plotRanges(z.GRanges)

```

# Using GRanges for genomic analysis
## Finding overlaps
```{r}
# load packages
library(GenomicFeatures)
library(GenomicRanges)
library(IRanges)
library(ERBS)

# load ESRRA ChIP data
data(HepG2)
data(GM12878)

browseVignettes("GenomicRanges")

# find binding sites common to both HepG2 and GM12878
?findOverlaps
# for each row in query, return overlapping row in subject
res = findOverlaps(HepG2, GM12878)
class(res)
res

# ranges from the query for which we found a hit in the subject
index = queryHits(res)
erbs = HepG2[index,]
erbs

# extract only the ranges
granges(erbs)
erbs
```

### Assessment
```{r}
library(ERBS)
data(HepG2)
data(GM12878)

start(HepG2)[17]

distanceToNearest(HepG2[17], GM12878)
GM12878[945]

# OR
# d = distanceToNearest(HepG2[17],GM12878)
# i = subjectHits(d)
# start(GM12878[i])

d = distanceToNearest(HepG2[17],GM12878)
mcols(d)$distance

d.all = distanceToNearest(HepG2, GM12878)
d.all.num = mcols(d.all)$distance
mean(d.all.num < 2000)
```

## Genes as GRanges
```{r}
## before video, run code from previous video AND these commands:
res = findOverlaps(HepG2,GM12878)
erbs = HepG2[queryHits(res)]
erbs = granges(erbs)

# define human genes
library(Homo.sapiens)
ghs = genes(Homo.sapiens)
ghs

# learn about the precede function (and related functions like follow)
?precede

# for each range in erbs, find the closest preceding range in ghs
index = precede(erbs, ghs)
ghs[index[1:3]]
erbs[1:3]    # note result is strand-aware
```

### Assessment
```{r}
genome(ghs)
length(ghs)

sort(table(seqnames(ghs)))
# which.max( table( seqnames( ghs ) ))

width(ghs)[1:50]
hist(width(ghs))
## the larger values are so much larger than the bulk of the data
## that we can barely see the frequencies of large values in the histogram 
mean(width(ghs))>median(width(ghs)) # a typical condition for right-skewed distributions

median(width(ghs))
```

## Finding the nearest gene
```{r}
# distance between binding sites and nearest preceding genes
distance(erbs, ghs[index])

# find transcription start site nearest to each binding site
tssgr = resize(ghs, 1)
tssgr

# distance between binding site and nearest TSS
d = distanceToNearest(erbs, tssgr)
queryHits(d)
dists = values(d)$distance
hist(dists, nc=100, xlim=c(0,100000))

index = subjectHits(d)[dists < 1000]
index
```

## Annotating genes and querying with "select"
```{r}
# use select function to query Homo sapiens database
tssgr[index,]
keytypes(Homo.sapiens)
keys = as.character(values(tssgr[index])$GENEID)
columns(Homo.sapiens)
res = select(Homo.sapiens, keys = keys,
             columns = c("SYMBOL", "GENENAME"), keytype="GENEID")
res[1:2,]
```

### Assessment
```{r}
library(ERBS)
data(HepG2)
data(GM12878)
res = findOverlaps(HepG2,GM12878)
erbs = HepG2[queryHits(res)]
erbs = granges(erbs)

erbs2= intersect(HepG2,GM12878)

erbs[erbs %over% erbs2]
erbs2[erbs %over% erbs2]

## first order them
erbs3 = erbs[order(erbs),]
##confirm same chr
all( seqnames(erbs2)==seqnames(erbs3) )
mean( start(erbs2)==start(erbs3) & end(erbs2)==end(erbs3) )
##the intersection should be smaller
all( width(erbs2) <= width(erbs3) )

# define human genes
library(Homo.sapiens)
ghs = genes(Homo.sapiens)
tss = resize(ghs,1)

# find transcription start site nearest to each binding site
tssgr = resize(ghs, 1)
start(tssgr["100113402"])

n = nearest(erbs[4], tss)
tss[n]
n.ID = mcols(tss)$GENEID[n]

keytypes(Homo.sapiens)
gene = as.character(n.ID) # specify with n.ID
select(Homo.sapiens,key=gene,column="SYMBOL",keytype="GENEID")
```

# Biostrings
## DNAString objects
```{r}
library(Biostrings)

# basics of DNAStrings
dna <- DNAString("TCGAGCAAT")    # define a DNAString
dna
length(dna)    # number of bases in a DNAString
# DNAString("JQX")    # error - invalid bases
# four bases (ACGT), the wild card or unknown base (N), and the dash representing a gap (-)
DNAString("NNNACGCGC-TTA-CGGGCTANN")    # valid sequence with unknowns and gaps
dna[4:6]    # extract a substring
as.character(dna)    # convert DNAString to character
```

## DNAStringSet objects
```{r}
# basics of DNAStringSets
set1 <- DNAStringSet(c("TCA", "AAATCG", "ACGTGCCTA", "CGCGCA", "GTT", "TCA"))    # define a DNAStringSet
set1
set1[2:3]    # extract subset of sequences
set1[[4]]    # extract one sequence as a single DNAString
length(set1)    # number of DNAstrings in set
width(set1)    # size of each DNAString
duplicated(set1)    # detect which sequences are duplicated
unique(set1)    # keep only unique sequences
sort(set1)
```

## Operations on DNAStrings
```{r}
dna_seq <- DNAString("ATCGCGCGCGGCTCTTTTAAAAAAACGCTACTACCATGTGTGTCTATC")

# analyze DNAStrings
letterFrequency(dna_seq, "A")    # count A in sequence
letterFrequency(dna_seq, "GC")    # count G or C in sequence
dinucleotideFrequency(dna_seq)    # frequencies of all dinucleotides
trinucleotideFrequency(dna_seq)    # frequencies of all trinucleotides

# convert DNAStrings
reverseComplement(dna_seq)    # find reverse complement
translate(dna_seq)    # amino acid translation
```

## Matching and counting with Biostrings
```{r}
# count and match on individual Biostrings
dna_seq <- DNAString("ATCGCGCGCGGCTCTTTTAAAAAAACGCTACTACCATGTGTGTCTATC")
dna_seq
countPattern("CG", dna_seq)    # pattern "CG" occurs 5 times
matchPattern("CG", dna_seq)    # locations of pattern "CG"
start(matchPattern("CG", dna_seq))    # start locations of the pattern
matchPattern("CTCTTTTAAAAAAACGCTACTACCATGTGT", dna_seq)    # match patterns of any length

# check for pattern and its reverse complement
countPattern("TAG", dna_seq)
countPattern(reverseComplement(DNAString("TAG")), dna_seq)

# count and match on sets of Biostrings
set2 <- DNAStringSet(c("AACCGGTTTCGA", "CATGCTGCTACA", "CGATCGCGCCGG", "TACAACCGTACA"))
set2
vcountPattern("CG", set2)    # CG counts for entire DNAStringSet
vmatchPattern("CG", set2)
vmatchPattern("CG", set2)[[1]]    # access matches for the first element of the DNAStringSet
```

### Assessment
```{r}
eco <- DNAString("GGTTTCACCGCCGGTAATGAAAAAGGCGAACTGGTGGTGCTTGGACGCAACGGTTCCGACTACTCTGCTGCGGTGCTGGCTGCCTGTTTACGCGCCGATTGTTGCGAGATTTGGACGGACGTTGACGGGGTCTATACCTGCGACCCGCGTCAGGTGCCCGATGCGAGGTTGTTGAAGTCGA")
eco

length(eco)
countPattern("ATG", eco)
start(matchPattern("ATG", eco))[1]
aa_full = translate(eco[17:length(eco)])
start(matchPattern("*", aa_full))

aa = aa_full[1:52]
length(aa)

letterFrequency(aa, "H")
letterFrequency(aa, "K")
letterFrequency(aa, "R")

letterFrequency(aa, "D")
letterFrequency(aa, "E")

letterFrequency(aa, "HKR") - letterFrequency(aa, "DE")
```

## Getting the sequence of regions
```{r}
library(ERBS)
data(HepG2)
HepG2

# load and inspect human reference genome
library(BSgenome.Hsapiens.UCSC.hg19)
Hsapiens

# extract chromosome 17 sequence
c17 = Hsapiens$chr17
c17

?getSeq
class(Hsapiens)
showMethods("getSeq")

# collection of DNA strings with ChIP-seq binding peaks
hepseq = getSeq(Hsapiens, HepG2)
length(HepG2)    # same number of sequences
width(HepG2)[1:5]    # widths match

# collection of shifted DNA strings with no relationship to binding sequences - essentially random
rhepseq = getSeq(Hsapiens, shift(HepG2, 2500))

# count occurrences of a motif in DNA sequences
mot = "TCAAGGTCA"
?vmatchPattern
vcountPattern(mot, hepseq)

# consider both forward matches and reverse complement matches 
sum(vcountPattern(mot, hepseq))    # forward pattern match
sum(vcountPattern(mot, reverseComplement(hepseq)))    # reverse pattern match

## compare motif occurrence in binding peak to random upstream sequences
# count of motifs in binding peaks
sum(vcountPattern(mot, hepseq)) +
    sum(vcountPattern(mot, reverseComplement(hepseq)))
# count of motifs in randomly selected regions of equal length
sum(vcountPattern(mot, rhepseq)) +
    sum(vcountPattern(mot, reverseComplement(rhepseq)))

# for real analysis, use MotifDb package, probabilistic binding packages like MEME and FIMO
```

### Assessment
```{r}
erbsseq = getSeq(Hsapiens, erbs)

GC.cont = letterFrequency(erbsseq, "GC")/width(erbsseq)
# OR
# GC.cont = rowSums(letterFrequency(erbsseq, "GC"))/width(erbsseq)
median(GC.cont)

rerbsseq = getSeq(Hsapiens, shift(erbs, 10000))
rGC.cont = letterFrequency(rerbsseq, "GC")/width(rerbsseq)
median(rGC.cont)
```

# Verified assessment
```{r}
library(Homo.sapiens)
g <- genes(Homo.sapiens)
g

genome(g)
length(g)
length(unique(seqlevels(g)))

# restrict to genes on chr21
chr21 <- keepSeqlevels(g, "chr21", pruning.mode = "coarse")
chr21
length(chr21)
max(width(chr21))
sum(strand(chr21) == "+")/length(chr21)

chr21seq = getSeq(Hsapiens, chr21)
GC.cont = rowSums(letterFrequency(chr21seq, "GC"))/width(chr21seq)
median(GC.cont)

chr21seq[5]
countPattern("ATG", chr21seq[[5]])
start(matchPattern("ATG", chr21seq[[5]]))
translate(chr21seq[[5]][32:width(chr21seq[5])])

snca <- genes(Homo.sapiens, filter=list(GENEID="6622"))
snca
tss = resize(snca, 1)
tss

snca_seq = getSeq(Hsapiens, snca)
countPattern("ACTGTGAA",snca_seq[[1]]) + countPattern("ACTGTGAA",reverseComplement(snca_seq[[1]]))
# OR
motif = "ACTGTGAA"
vcountPattern(motif, snca_seq) + vcountPattern(reverseComplement(DNAString(motif)), snca_seq)

# Load the GRanges corresponding to human genes from the Homo.sapiens package
g <- genes(Homo.sapiens)

# Load the ESRRA binding site GRanges in GM12878 cells
data(GM12878)

g[100]
p = flank(g[100], 2000)
p.all=flank(g, 2000)

overlap = findOverlaps(p.all,GM12878)
length(unique(subjectHits(overlap)))
length(unique(queryHits(overlap)))

shift.p.all = shift(p.all,10000)
shift.overlap = findOverlaps(shift.p.all,GM12878)
length(unique(subjectHits(shift.overlap)))

length(unique(subjectHits(overlap)))/length(unique(subjectHits(shift.overlap)))
```


