---
title: "Section 3: Management of genome-scale data with Bioconductor"
author: "Haeun Lee"
output: html_document
date: "2026-12-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
BiocManager::install(c("Biobase",
                       "GEOquery",
                       "genomicsclass/GSE5859Subset",
                       "affy",
                       "hgu95acdf",
                       "genefilter",
                       "parathyroidSE",
                       "airway",
                       "pasillaBamSubset",
                       "Rsamtools",
                       "GenomicAlignments",
                       "ArrayExpress",
                       "NGScopyData",
                       "AnnotationDbi"))
```

# Organizing microarray data with ExpressionSet
## The ExpressionSet class
```{r}
library(Biobase)
library(GEOquery)

geoq <- getGEO("GSE9514")    # download a microarray dataset from GEO
names(geoq)    
e <- geoq[[1]]    # extract ExpressionSet
e

# exprs gives matrix of microarray values
dim(e)    # number of features and samples in ExpressionSet
ncol(e)
nrow(e)

exprs(e)[1:3,1:3]
head(exprs(e))[,1]    # first column
exprs(e)[1,]    # first row
exprs(e)["10000_at",]    # can also index by name
rownames(e)[1]    # row names are probe sets
dim(exprs(e))    # rows are features, columns are samples

# pData gives phenotype data (sample information)
pData(e)[1:3,1:6]
names(pData(e))
pData(e)$characteristics_ch1    # column in GEO to describe experimental state/condition
as.numeric(factor(pData(e)$characteristics_ch1))    # help see replicates of each state
dim(pData(e))    # rows of pData correspond to columns of exprs
dim(e)

# fData gives feature data (probe information)
fData(e)[1:3,1:3]
dim(fData(e))    # rows of fData correspond to rows of exprs
names(fData(e))
head(fData(e)$"Gene Symbol")
head(rownames(e))

# additional annotation tied to ExpressionSet
experimentData(e)
annotation(e)
```

### Assessment
```{r}
library(GSE5859Subset)
data(GSE5859Subset)
dim(geneExpression) # assayData
dim(sampleInfo) # phenoData
dim(geneAnnotation) # featureData

# rows sampleInfo = columns geneExpression
  # rownames() of phenoData = colnames() of assayData
# rows geneAnnotation = rows geneExpression
  # rownames() of featureData = rownames() of assayData
identical(colnames(geneExpression), sampleInfo$filename)
identical(rownames(geneExpression), geneAnnotation$PROBEID)

pd <- AnnotatedDataFrame(sampleInfo)
rownames(pd) <- pd$filename
pd <- pd[, varLabels(pd) != "filename"] ## redundant

pData(pd)["GSM136530.CEL.gz", "date"]
varLabels(pd)

fd <- AnnotatedDataFrame(geneAnnotation)
rownames(fd) <- fd$PROBEID
fd <- fd[, varLabels(fd) != "PROBEID"] ## redundant
pData(fd)["204810_s_at", "CHR"]

eset <- ExpressionSet(geneExpression,pd,fd)
eset

ind1 <- which(featureData(eset)$CHR == "chrY")
ind2 <- pData(eset)$group == 1
femaleY <- colMeans(exprs(eset)[ind1, ind2]) 
maleY <- colMeans(exprs(eset)[ind1, !ind2]) 
boxplot(maleY, femaleY)
median(maleY) - median(femaleY)

dim(exprs(eset))
dim(pData(eset))
dim(fData(eset))

annotation(eset) <- "hgfocus"

eset[1:10, 1:5] # first 10 feature and first 5 samples
```

## Reading microarray raw data: single-color arrays
```{r}
library(affy)
library(here)

# base directory
basedir <- here("rawdata-master", "celfiles")

# read sample info
tab_file <- file.path(basedir, "sampleinfo.txt")
tab <- read.delim(tab_file, check.names=FALSE, as.is=TRUE)
rownames(tab) <- tab$filenames

# Read and process Affy batch with full paths
ab <- ReadAffy(filenames = file.path(basedir, tab$filenames), phenoData = tab)

# inspect dimensions
dim(pm(ab))    # probe-level data; pm: perfect match probe-level intensity, experimental data
dim(pData(ab)) # sample-level metadata

rownames(ab)[1:10]
colnames(pm(ab))
annotation(ab)

# preprocess to gene-level using RMA (probe-level -> gene-level data)
e <- rma(ab)

# OR one-step processing (read and process data to gene-level)
ejust <- justRMA(filenames = tab$filenames, phenoData = tab,
                 celfile.path = basedir)
dim(ejust)
```

## Reading microarray raw data: Agilent two-color arrays
```{r}
 # datadir defined in previous video
library(limma)
library(rafalib)

# base directory relative to project root
basedir <- here("rawdata-master", "agilent")

# full path to the target file
target_file <- file.path(basedir, "TargetBeta7.txt")

# read targets safely
targets <- readTargets(target_file)

# source: which imaging software was used to produce file
# read Red/Green intensities
RG <- read.maimages(targets$FileName, source="genepix", path=basedir)

# M: log ration, A: Average of log
# transformed
MA <- MA.RG(RG,bc.method="none")

# dimensions
dim(RG$R)
dim(RG$G)
dim(MA$M)
dim(MA$A)

# MA plot for first sample
plot(MA$A[,1], MA$M[,1])

# microarray image
mypar(1,1)
imageplot(MA$M[,2], RG$printer, zlim=c(-3,3))
dev.off()
```

### Assessment
```{r}
# Q1
library(hgu95acdf)

# set base directory (absolute path; change if needed)
basedir <- "/Users/iha-eun/Desktop/HarvardX/01. Introduction to Bioconductor/rawdata-master/celfiles"

# read sample info
tab <- read.delim(file.path(basedir, "sampleinfo.txt"),
                  check.names = FALSE, as.is = TRUE)

rownames(tab) <- tab$filenames
tab["1521a99hpp_av06.CEL.gz","36311_at"]

# Q2
ab <- ReadAffy(filenames = file.path(basedir, tab[,1]), phenoData=tab)
feature.ID <- probeNames(ab) # extract feature ID
sum(feature.ID == "36311_at")

# probes are often used for each gene (feature)
length(featureNames(ab)) # number of genes
length(probeNames(ab)) #  number of probes

# Q3
# ab_sub <- ab[c("1532a99hpp_av04.CEL.gz","1532b99hpp_av04.CEL.gz")]
# mat <- pData(ab_sub)[,"36085_at"]

# Explanation
## identify columns w/ samples of interest
samps <- c("1532a99hpp_av04.CEL.gz", "1532b99hpp_av04.CEL.gz")
ind <- which(colnames(ab) %in% samps)
ab[, ind]
ab_subb <- ab[, ind]

## extract probe intensities for probe set of interest
pid <- "36085_at"
# pm() get the data from assayData
matt <- pm(ab_subb)[probeNames(ab_subb) == pid, ] # pm; extracts the perfect match intensities

## determine expected concentrations
# pData() get info from PhenoData
conc <- pData(ab)[ind, pid]

## compute probe-level log2 fold changes
lfc <- log2(matt[, 2] / matt[, 1])

## make the plots
boxplot(lfc, ylim = c(-0.5, 1.5))
abline(h = log2(conc[2] / conc[1]), lty = 2, col = 'blue') #intended log fold
abline(h = 0)
```

# Organizing NGS data with SummarizedExperment
## The SummarizedExperiment class
```{r}
library(parathyroidSE)
data(parathyroidGenesSE)
se <- parathyroidGenesSE
se

# assay contains results of the assay
dim(se)
assay(se)[1:3,1:3]
dim(assay(se))    # rows = features (ranges), columns = samples

# colData contains sample information (e.g. gene or transcript IDs, etc.)
colData(se)[1:3,1:6]
dim(colData(se))
names(colData(se))
colData(se)$treatment
as.numeric(colData(se)$treatment)

# rowRanges contains feature information (e.g. genes, transcripts, exons, etc.)
# genomic ranges of interest
rowRanges(se)[1]
class(rowRanges(se))
length(rowRanges(se))    # number of genes
length(rowRanges(se)[[1]])    # number of exons for first gene
head(rownames(se))
metadata(rowRanges(se)) # how this GRanges list was constructed

# additional metadata, including sample information
metadata(se)$MIAME # MIAME; Minimal Information About Microarray Experiment
abstract(metadata(se)$MIAME)
```

## Assessment
```{r}
library(airway)
data(airway)
airway    # SummarizedExperiment object

metadata(airway)
length(rowRanges(airway))
nrow(airway)
length(colData(se))
ncol(airway)

which(colData(airway)$Run == "SRR1039509")
colData(airway)$cell[2]
colData(airway)["SRR1039509", "cell"]    # inspect column names to figure out which refers to cell line: "cell"

names(colData(airway))    # look for name related to dexamethasone: "dex"

mean(assay(airway)[,1])
mean(assay(airway)[,2])
mean(assay(airway)[,3])
mean(assay(airway)[,4])
mean(assay(airway)[,5])
mean(assay(airway)[,6])
mean(assay(airway)[,7])
mean(assay(airway)[,8])
rownames(colData(airway))[which.min(colData(airway)$avgLength)]    # look for column name related to read length: avgLength

dim(assay(airway))
length(rowRanges(airway)) # number of gene
rowRanges(airway)[100]
length(rowRanges(airway)[[100]]) # number of exon in 100th gene
seqnames(rowRanges(airway)[[100]])
range(rowRanges(airway)[[100]])
width(range(rowRanges(airway)[[100]]))
tss = resize(range(rowRanges(airway)[[100]]),1)
tss

ENSG <- airway["ENSG00000103196"]
mean(ENSG@assays@data@listData[["counts"]])
mean(assay(airway)["ENSG00000103196", ])

trt.mean <- mean(ENSG@assays@data@listData[["counts"]][which(colData(airway)$dex == "trt")])
trt <- mean(assay(airway)["ENSG00000103196", which(colData(airway)$dex == "trt")])
untrt.mean <- mean(ENSG@assays@data@listData[["counts"]][which(colData(airway)$dex == "untrt")])
untrt <- mean(assay(airway)["ENSG00000103196", which(colData(airway)$dex == "untrt")])
log2(trt.mean/untrt.mean)
```

# Importing NGS data in R
```{r}
# How to import NGS data using Rsamtools
library(pasillaBamSubset)
library(Rsamtools)
filename <- untreated1_chr4()

bf <- BamFile(filename)

seqinfo(bf)
sl <- seqlengths(bf)

quickBamFlagSummary(bf)

# Specifying: what and which
## what - what kind of information to extract?
## which - which ranges of alignments to extract?

gr <- GRanges("chr4", IRanges(1, sl["chr4"]))
countBam(bf, param = ScanBamParam(which = gr))

reads <- scanBam(BamFile(filename, yieldSize = 5))

# Examining the output of scanBam
class(reads)
names(reads[[1]])
reads[[1]]$pos    # the aligned start position
reads[[1]]$rname    # the chromosome
reads[[1]]$strand    # the strand
reads[[1]]$qwidth    # the width of the read
reads[[1]]$seq    # the sequence of the read

gr <- GRanges("chr4", IRanges(500000, 700000))
reads <- scanBam(bf, param = ScanBamParam(what = c("pos","strand"), which = gr))

# The GenomicAlignments package
library(GenomicAlignments)
ga <- readGAlignments(bf)
length(ga)

granges(ga[1])

gr <- GRanges("chr4", IRanges(700000, 800000))
(fo <- findOverlaps(ga, gr))    # which reads over this range
countOverlaps(gr, ga)    # count overlaps of range with the reads
table(ga %over% gr)    # logical vector of read overlaps with the range
```

### Assessment
```{r}
# pasillaBamSubset; BAM file containing a subset of the reads from a single-end RNA-seq experiment in Drosophila melanogaster (fruit fly), aligning to chromosome 4
library(pasillaBamSubset)
library(Rsamtools)
filename <- untreated1_chr4()
bf <- BamFile(untreated1_chr4())

gr <- GRanges("chr4",IRanges(440000, 470000))
countBam(bf, param=ScanBamParam(which = gr))
# or, alternatively:
reads <- scanBam(bf, param = ScanBamParam(what = "pos", which = gr))
length(reads[[1]]$pos)

GC <- scanBam(bf, param=ScanBamParam(what = "seq", which = gr))
GC.prop <- letterFrequency(GC[["chr4:440000-470000"]][["seq"]],"CG") / GC[["chr4:440000-470000"]][["seq"]]@ranges@width
mean(GC.prop)
# or, alternatively:
mean(letterFrequency(GC[[1]]$seq, "GC", as.prob=TRUE))

library(GenomicAlignments)
ga <- readGAlignments(BamFile(filename))

# "spiky" look of this histogram = genes are spread out on the chromosome, some genes have much higher gene expression
hist(start(ga), breaks=100)

BiocManager::install("TxDb.Dmelanogaster.UCSC.dm3.ensGene")
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
g <- genes(TxDb.Dmelanogaster.UCSC.dm3.ensGene)    # GRanges corresponding to genes

g2 <- g[g %over% GRanges("chr4", IRanges(200000, 300000))]
countOverlaps(g2["FBgn0039890"],ga)

```

## Creating a count table from a BAM file

Count sequence reads
Count matrix
- Row: Genomic ranges (~ featureData in ExpressionSet)
- Column: Samples (~ phenoData in ExpressionSet)
```{r}
# libraries with BAM files and annotation
library(pasillaBamSubset)
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
txdb <- TxDb.Dmelanogaster.UCSC.dm3.ensGene # txdb; transcript databases

# counting matrix
grl <- exonsBy(txdb, by="gene")    # make GRangesList of exons for each gene
grl[100]    # GRangesList of exons for 100th gene
grl[[100]]    # GRanges with exons of 100th gene
grl[[100]][1]    # first exon of 100th gene

# paths to BAM files
fl1 <- untreated1_chr4()
fl2 <- untreated3_chr4()

# libraries for importing BAM files
library(Rsamtools)
library(GenomicRanges)
library(GenomicAlignments)

# specify files with BamFileList
fls <- BamFileList(c(fl1, fl2))
names(fls) <- c("first","second")

# find reads that overlap exons
# count the reads (fls) which falls into features (grl)
so1 <- summarizeOverlaps(features=grl,
                         reads=fls,
                         ignore.strand=TRUE)
so1 # class: RangedSummarizedExperiment, count matrix available @assay

# examine count matrix
head(assay(so1))
colSums(assay(so1))

# examine rest of SummarizedExperiment components
rowRanges(so1) # features, genes
colData(so1) # sample
colData(so1)$sample <- c("one","two")    # add sample information
colData(so1)
metadata(rowRanges(so1)) 

# exploratory data analysis of counts
x <- assay(so1)[,1]
hist(x[x > 0], col="grey")
hist(x[x > 0 & x < 10000], col="grey")
plot(assay(so1) + 1, log="xy") # 2 samples are highly correlated, if one sample has high counts, other also does

# count second file as paired-end reads
# ?untreated3_chr4
# ?summarizeOverlaps
fls <- BamFileList(fl2)
so2 <- summarizeOverlaps(features=grl,
                         reads=fls,
                         ignore.strand=TRUE,
                         singleEnd=FALSE, 
                         fragments=TRUE) # if only one from paired reads maps, also count
colSums(assay(so2)) # paired reads counted once
colSums(assay(so1)) # paired reads counted twice (total count doubled)

# show there are half as many reads in so2 as so1
plot(assay(so1)[,2], assay(so2)[,1], xlim=c(0,5000), ylim=c(0,5000),
     xlab="single end counting", ylab="paired end counting")
abline(0,1)
abline(0,.5)
```

### Assessment
```{r}
library(pasillaBamSubset)
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
txdb <- TxDb.Dmelanogaster.UCSC.dm3.ensGene

# "FB..." = gene name
g <- genes(txdb)
g <- g[seqnames(g) == "chr4"]

# Create the GRangesList of exons by gene
# subset to the same genes as in g
grl <- exonsBy(txdb, by="gene")
grl <- grl[names(g)]

all.equal(names(g), names(grl))

library(Rsamtools)
bf <- BamFile(untreated1_chr4())

# reads which fall in introns may not be from the fully mature mRNA
so1 <- summarizeOverlaps(g, bf, ignore.strand=TRUE)
# GRangesList feature set (so2) gives a lower count value
# because this method of counting does not include reads which fall in introns
so2 <- summarizeOverlaps(grl, bf, ignore.strand=TRUE)

colSums(assay(so1)) # include introns, higher count
colSums(assay(so2)) # exon only

zero.index = which(so1@assays@data@listData[["counts"]] == "0")
so1 <- so1[-zero.index,]
so2 <- so2[-zero.index,]

x <- assay(so1)[,1]
y <- assay(so2)[,1]
par(mfrow=c(2,1))
hist(x[x > 0 & x < 5000], col="blue")
hist(y[y > 0 & y < 5000], col="red")

plot(assay(so1)[,1], assay(so2)[,1], xlim=c(0,5000), ylim=c(0,5000),
     xlab="GRanges object", ylab="GRangesList object")
abline(0,0)
abline(0,1)

mean(so2@assays@data@listData[["counts"]]/so1@assays@data@listData[["counts"]])

#OR
g.so <- summarizeOverlaps(g, bf, ignore.strand = TRUE)
grl.so <- summarizeOverlaps(grl, bf, ignore.strand = TRUE)

# this generates a warning about applying log to 0s
plot(assay(g.so), assay(grl.so), log = "xy")
abline(0, 1)

# adding a pseudocount prevents warnings for 0s
plot(assay(g.so) + 1, assay(grl.so) + 1, log = "xy")
abline(0, 1)

ratio <- assay(grl.so) / assay(g.so)
mean(ratio[assay(g.so) > 0])

# scales each column of the count table such that we get the number of reads expected if the sample were sequenced to have 1 million reads mapping to genes on chromosome 4
# reads per million (RPM) OR fragments per million (FPM)
assay(grl.so)/sum(assay(grl.so))*1e6

# OR
count <- assay(grl.so)
# here, we can just use sum() because there is only one column
# otherwise we would use: sweep(count, 2, colSums(count), "/")
fpm <- (count / sum(count)) * 1e6
fpm[1]

library(data.table)
grl.names <- grl@partitioning@NAMES
unlist <- unlist(grl)
split <- split(grl,grl.names)

# Assuming 'simplelist' is your SimpleList of GRangesList objects
# Define a function or use an anonymous function to process each GRangesList
process_grl <- function(grl_element) {
  # Inside this function, 'grl_element' is a single GRangesList
  # You can now operate on this GRangesList, for example, unlist it
  # or apply further GRangesList-specific operations.
  return(width(reduce(grl_element))) # Example operation: convert each GRangesList to a single GRanges object
}

grl.sub <- lapply(split,process_grl)
sum(grl.sub$FBgn0002521)

# FPKM, the number of fragments per kilobase of exonic basepairs, per million mapped reads
# FPM / number of basepairs in the exons of each gene * 1000
4.275607e+03/2399*1000

# OR
ebp <- sum(width(reduce(grl)))
count <- assay(grl.so)
fpm <- (count / sum(count)) * 1e6
fpkm <- (fpm / ebp) * 1e3
fpkm[1]
```
Note that there are more robust ways of estimating FPM and FPKM. For example, using a more robust estimate of the sequencing depth, and taking into account multiple transcript isoforms, which might be expressed at different levels. Here, we have demonstrated the simplest kind of sequencing depth and gene length normalization introduced for RNA-seq analysis.

# Public data repositories
## GEOquery: accessing 1000s of microarray experiments at NCBI
```{r}
library(GEOquery)
glioMA = getGEO("GSE78703")[[1]]
glioMA

names(pData(glioMA))    # variable names
glioMA$molecule_ch1    # molecule being assayed (RNA)
table(glioMA$`treated with:ch1`, glioMA$`cell type:ch1`)    # experimental variables
```

## Array-oriented annotation
```{r}
#GPL: different platforms are listed
library(hgu133plus2.db)
hgu133plus2.db

#library(hgu133plus2probe)
#head(hgu133plus2probe)
#dim(hgu133plus2probe)

# PROBID = probe set information
# 1 probe set contains several 25mers
# Map: cytoband
select(hgu133plus2.db, keytype="PROBEID", 
       columns=c("SYMBOL", "GENENAME", "PATH", "MAP"), keys="1007_s_at")
```

### Assessment
```{r}
columns(hgu133plus2.db)
EGFR <- select(hgu133plus2.db, keytype="SYMBOL", 
       columns="PROBEID", keys="EGFR")
unique(EGFR$PROBEID)
# OR
nrow(select(hgu133plus2.db, keys="EGFR", columns="PROBEID", keytype="SYMBOL"))

library(GO.db)
columns(GO.db)

select(GO.db, keys="glial cell proliferation", columns=c("DEFINITION","GOID","ONTOLOGY","TERM"), keytype="TERM")

select(hgu133plus2.db,keytype="GO",keys="GO:0014009",columns="PROBEID")

key = select(GO.db, keys="glial cell proliferation", keytype="TERM", columns="GOID")$GOID
nrow(select(hgu133plus2.db, keys=key, columns=c("PROBEID", "SYMBOL"), keytype="GO"))

```

# Verified assessment
```{r}
library(GEOquery)
library(NGScopyData)
library(Rsamtools)
library(GenomicAlignments)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(AnnotationDbi)
```

```{r}
lstem <- getGEO("GSE3725")[[1]]

ncol(lstem)
nrow(lstem)

lstem <- lstem[, !grepl("^GMP expressing", pData(lstem)$title)]
titles <- as.character(pData(lstem)$title)
cell_type <- gsub(".*\\((.*?)( enriched)?\\).*", "\\1", titles)

# add cell_type column to pData
pData(lstem)$cell_type <- factor(cell_type)

sum(pData(lstem)$cell_type == "L-GMP")

index = which(fData(lstem)$ID == "1421579_at")
fData(lstem)$"Gene Symbol"[index]
# fData(lstem["1421579_at",])$"Gene Symbol"

mean(exprs(lstem)["1421579_at",])
# mean(exprs(lstem["1421579_at",]))

lstem.L.GMP <- lstem[,pData(lstem)$cell_type == "L-GMP"]
mean(exprs(lstem.L.GMP)["1421579_at",])
# mean(exprs(lstem["1421579_at", pData(lstem)$cell_type == "L-GMP"]))

```

```{r}
library(NGScopyData)
tps_27 <- tps_27.chr6()$bamFpath    # path to BAM file

gr <- GRanges("chr6",IRanges(260000, 360000))
bf <- BamFile(tps_27)
reads <- scanBam(bf, param = ScanBamParam(what = c("pos","strand"), which = gr))[[1]]

sum(reads$strand == "+") / length(reads$strand)

hist(reads$pos)

library(GenomicAlignments)
ga <- readGAlignments(bf)
length(ga)
# seqnames(x): Get or set the name of the reference sequence for each alignment
# strand(x):  Get or set the strand for each alignment
# qwidth(x): (i.e. the length of the query sequence that is stored in the corresponding SAM/BAM record)

mean(qwidth(ga))
hist(ga@start)

# load GRanges with human genes
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
g <- genes(txdb)

# subset to genes on chr6
g_chr6 <- g[seqnames(g)=="chr6"]

se <- summarizeOverlaps(g_chr6, bf, ignore.strand=TRUE)
sum(assays(se)$count)
sum(assays(se)$count == "0")
# OR
se <- summarizeOverlaps(features=g_chr6,
                          reads=tps_27,
                          ignore.strand=TRUE)
sum(assay(se))
sum(assay(se) == 0)

count <- assay(se)
fpm <- (count / sum(count)) * 1e6
max(fpm)

identical(names(g_chr6), rownames(assay(se)))

fpkm <- fpm/width(ranges(g_chr6))*1e3
median(fpkm)
mean(fpkm)
sum(fpkm > 1000)

index = which(fpkm > 1000)
top_entrez <- rowRanges(se)$"gene_id"[index]
# top_entrez = rownames(fpkm)[fpkm > 1000]
AnnotationDbi::select(org.Hs.eg.db, keys=top_entrez, keytype = "ENTREZID", columns = "SYMBOL")
```


