---
title: "Section 4: Genomic annotation with Bioconductor"
author: "Haeun Lee"
output: html_document
date: "2026-01-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Bioconductor packages
BiocManager::install(c("BSgenome",
                       "BSgenome.Hsapiens.UCSC.hg19.masked",
                       "org.Hs.eg.db",
                       "ensembldb",
                       "EnsDb.Hsapiens.v75",
                       "AnnotationHub",
                       "rtracklayer",
                       "TxDb.Hsapiens.UCSC.hg38.knownGene",
                       "KEGGREST",
                       "rols",
                       "GSEABase"))

# CRAN packages
# install.packages(c("R.utils",
#                    "png",
#                    "DT"))
```

# Annotation of genes and transcripts
## Reference sequence packages for model organisms
```{r}
library(BSgenome)
library(Biostrings)
# reference genomes
ag = available.genomes()
grep("Scerev", ag, value=TRUE)
grep("Hsap", ag, value=TRUE)

# inspect the human genome
library(BSgenome.Hsapiens.UCSC.hg19)
Hsapiens
length(Hsapiens)
class(Hsapiens)
methods(class="BSgenome")

# inspect human genome
Hsapiens$chrX
substr(Hsapiens$chrX, 5e6, 5.1e6)
nchar(Hsapiens$chrY)
nchar(Hsapiens[[24]])
```
In Biostrings, a pile of masks can be put on top of a sequence. A pile of masks is represented by a MaskCollection object and the sequence by an XString object. A MaskedXString object is the result of bundling them together in a single object.

### Assessment

Genomic sequence "mask" structures have been defined to isolate ambiguous, unmappable, and low-complexity segments of genomes so that sequence analysis research can be targeted to reflect current knowledge of sequence regions that are more likely to be functionally informative.
```{r}
library(BSgenome)
grep("mask", grep("Drerio", available.genomes(), value=TRUE), invert=TRUE, value=TRUE) # exclude masked

library(BSgenome.Hsapiens.UCSC.hg19.masked)
# original sequence is always stored unmodified in a MaskedXString object
c17m = BSgenome.Hsapiens.UCSC.hg19.masked$chr17
c17 = BSgenome.Hsapiens.UCSC.hg19$chr17 # contain 'unmasked' from c17m only

class(c17m)

c22m = BSgenome.Hsapiens.UCSC.hg19.masked$chr22
print(c22m)
maskedratio(c22m)*100
# OR
# the masks of assembly gaps, aka “the AGAPS masks”;
# the masks of intra-contig ambiguities, aka “the AMB masks”;
# the masks of repeat regions that were determined by the RepeatMasker software, aka “the RM masks”;
# the masks of repeat regions that were determined by the Tandem Repeats Finder software (where only repeats with period less than or equal to 12 were kept), aka “the TRF masks”
round(100*sum(width(masks(c22m)$AGAPS))/length(c22m),0)
```

## Packages for gene and transcript catalogs

Information about reference genome regions like genes, transcripts and exons can also be found in annotation packages. One major source of reference genome structure annotation is the UCSC Genome Browser.
```{r}
# import TxDb transcript database
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb = TxDb.Hsapiens.UCSC.hg19.knownGene  # abbreviate
class(txdb) # GenomicFeatures
methods(class="TxDb")

# extract and inspect genes from TxDb
# returns a GRanges. Each range corresponds to a gene.
genes(txdb)
# GRanges and IRanges package tools can be used
# counts the number of genes on each strand
table(strand(genes(txdb)))
# summary statistics describing the length in base pairs of genes in the human genome
summary(width(genes(txdb)))

# inspect largest gene in genome
# which row in the GRanges represents the largest gene
id = which.max(width(genes(txdb)))
# gene_id metadata column lists the ENTREZ ID for a given gene
genes(txdb)[id]
library(org.Hs.eg.db)
anno =  org.Hs.eg.db
# ENTREZ IDs can be converted into meaningful information like gene symbol and name
select(org.Hs.eg.db, keys="286297", keytype = "ENTREZID", columns = c("SYMBOL", "GENENAME"))

# compare total size of exons to total size of genes
# TxDb also includes a dataset for all known exons
ex = exons(txdb)
# fraction of bases in gene regions expressed as exons
# some exons overlap, reduce() condenses exons into non-overlapping ranges so that every base in an exon is only counted once
rex = reduce(ex)
ex_width = sum(width(rex))    # bases in exons
gene_width = sum(width(genes(txdb)))    # bases in genes
ex_width/gene_width # 6.38% of bases in genes are exons, coding fraction of gene regions
```

## ensembldb, EnsDb: annotation from EMBL

There are different annotation sources available: UCSC, Ensemble, etc
-> Different resource gives different information
```{r}
# inspect data available from Ensembl
library(ensembldb)
library(EnsDb.Hsapiens.v75)
names(listTables(EnsDb.Hsapiens.v75))

# extract Ensembl transcripts
edb = EnsDb.Hsapiens.v75  # abbreviate
txs <- transcripts(edb, filter = GeneNameFilter("ZBTB16"),
                   columns = c("protein_id", "uniprot_id", "tx_biotype"))
txs

# compare Ensembl and UCSC transcripts
alltx = transcripts(edb)    # Ensembl is larger
utx = transcripts(txdb)    # UCSC is smaller

# table of biological types of transcripts
table(alltx$tx_biotype)
```

### Assessment
```{r}
BiocManager::install("Gviz")
library(devtools)
install_github("genomicsclass/ph525x")
library(ph525x)
stopifnot(packageVersion("ph525x") >= "0.0.16") # do over if fail 

# The linear structures are transcripts, composed of untranslated regions at the ends, exons (small yellow rectangles), and introns (arrows between exons). The collection of transcripts makes up a "gene model".
modPlot("ESR1", useGeneSym=FALSE, collapse=FALSE)

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb = TxDb.Hsapiens.UCSC.hg19.knownGene  # abbreviate

# gene_id = 2099

txs.2099 <- transcriptsBy(txdb, by="gene")$"2099"
length(txs.2099)
```

## AnnotationHub, liftOver and rtracklayer
```{r}
# install.packages("remotes")

# remotes::install_version("dbplyr", version = "2.3.4")
# remotes::install_version("dplyr",  version = "1.1.2")
# remotes::install_version("DBI", version = "1.2.3")

library(AnnotationHub)
ah <- AnnotationHub()
ah

methods(class="AnnotationHub")

# lists the number of unique species which currently have records in AnnotationHub
length(unique(ah$species))

# filters records that match specific fields
ah_human <- subset(ah, species == "Homo sapiens")
ah_human

# searches for records that match a desired string in any of their fields. Unlike subset(), you do not need to specify which field to search.
query(ah, "HepG2")
# multiple terms at once
query(ah, c("HepG2", "H3K4me3"))
hepg2 <- query(ah, "HepG2")
hepg2_h3k4me3 <- query(hepg2, c("H3K4me3"))
hepg2_h3k4me3
# annotation metadata
hepg2_h3k4me3$tags

# # search records interactively by using display on a query
# display(query(ah, "HepG2"))

# To load a record from AnnotationHub, you need its ID. These commands find the ID number of the HepG2 H3K4me3 broadPeak file
e118_broadpeak <- query(hepg2_h3k4me3, c("E118", "broadPeak"))
id <- e118_broadpeak$ah_id
id

hepg2_h3k4me3_broad <- ah[["AH29728"]]
hepg2_h3k4me3_broad
# store the ID as a variable and use that variable for indexing
alt_format <- ah[[id]]
identical(hepg2_h3k4me3_broad, alt_format)
```

### Assessment
```{r}
library(AnnotationHub)
ah = AnnotationHub()
mah = mcols(ah)
names(mah)
sort(table(mah$species), decreasing=TRUE)[1:10]

# files that address binding of CTCF to DNA from HepG2 cells
HepG2.CTCF = query(query(ah, "HepG2"), "CTCF")
HepG2.CTCF
```
## liftOver: Translating between reference builds
```{r}
# learn about liftOver from rtracklayer
library(rtracklayer)
?liftOver

# Suppose we want to convert chromosome 1 gene locations from hg38 to hg19
# get the GRanges for those genes from the TxDb for hg38
# chromosome 1 gene locations in hg38
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
tx38 <- TxDb.Hsapiens.UCSC.hg38.knownGene
# restricts the TxDb to features on chromosome 1
seqlevels(tx38, pruning.mode="coarse") = "chr1"
# acquires the GRanges for chromosome 1 genes
g1_38 <- genes(tx38)

# download the hg38 to hg19 chain file
library(AnnotationHub)
ah <- AnnotationHub()
# chain file that specifies how features match across genome builds
# searches AnnotationHub for chain files that relate hg19 and hg38
ah.chain <- subset(ah, rdataclass == "ChainFile" & species == "Homo sapiens")
query(ah.chain, c("hg19", "hg38"))
# choose the record titled hg38toHg19.over.chain.gz.
ch <- ah[["AH14108"]]

# perform the liftOver
# GRanges as the first argument and the chain file as the second argument
g1_19L <- liftOver(g1_38, ch)
g1_19L
```

### Assessment
```{r}
# download.file("http://hgdownload.cse.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz", "hg19ToHg38.over.chain.gz")
# library(R.utils)
# gunzip("hg19ToHg38.over.chain.gz")

library(ERBS)
data(HepG2)
library(rtracklayer)
# convert the HepG2 binding addresses to hg38
ch = import.chain("hg19ToHg38.over.chain") 
nHepG2 = liftOver(HepG2, ch)
nHepG2

start(HepG2)[1] - start(nHepG2)[1]
```

## Data import and export with rtracklayer
```{r}
library(devtools)
install_github("genomicsclass/ERBS")    # install ERBS package

# finds the path to the first file in the extdata directory of the ERBS package on your machine and assigns that file path to f1
# narrowPeak file containing ER binding data from ENCODE
f1 = dir(system.file("extdata",package="ERBS"), full=TRUE)[1]    # access data
# follows a similar format to BED files with 4 additional columns specific to narrowPeak data
# raw text that is tab-separated into columns
readLines(f1, 4)    # look at a few lines

# input file and a format argument that specifies how to parse that file into an object
library(rtracklayer)
imp = import(f1, format="bedGraph")    # import as bedGraph format
# extra columns in the narrowPeak format are not named
# can add metadata column names with names(mcols(imp))
imp

# set genome explicitly to ensure your data is processed correctly downstream
# genome of the GRanges is NA by default
genome(imp)    # genome identifier tag not set, but you should set it
genome(imp) = "hg19"    # set genome
genome(imp)

# export() takes an object and writes an output file in a common genomic file format
# GRanges gets converted into a BED format file
export(imp, "demoex.bed")    # export as BED format  
cat(readLines("demoex.bed", n=5), sep="\n")    # check output file
```

```{r}
library(rtracklayer)
data(targets) # microRNA target sites

# It is not typical to add metadata to data.frame instances, as one has to fill a column with relevant information -> no reference genome information available in targets
class(targets)
summary(targets)
targets

library(GenomicRanges)
mtar <- with(targets,
GRanges(chrom, IRanges(start,end), strand=strand,
targets=target, mirname=name))

cat(export(mtar[1:5], format="bed"), sep="\n")
cat("\n")
cat(export(mtar[1:5], format="gff3"), sep="\n")

mtar[1]$mirname # name of microRNA
# OR in GFF3 format
```
# Annotation tools for systems biology
## OrgDb: unified organism-specific annotation for systems biology
Packages named org.*.eg.db collect information at the gene level with links to location, protein product identifiers, KEGG pathway and GO terms, PMIDs of papers mentioning genes, and to identifiers for other annotation resources.
```{r}
# load human OrgDb and inspect available keys
# org: organism level annotation
# Hs: Homo Sapiens
# eg: Entrez gene as resource
# db: SQLite based package
library(org.Hs.eg.db)
org.Hs.eg.db
keytypes(org.Hs.eg.db) # columns() gives same answer

# load GO.db and inspect available terms
library(GO.db)
allterms = keys(GO.db, keytype="TERM")
allterms[1:5]

# find GOID (gene ontology tag) for ribosome biogenesis
select(GO.db,  keys = "ribosome biogenesis", keytype="TERM", columns="GOID")

# find symbols for genes involved in ribosome biogenesis 
select(org.Hs.eg.db, keys = "GO:0042254", keytype="GO", columns="SYMBOL")

# you can pull out multiple columns at once
select(org.Hs.eg.db, keys = "GO:0042254", keytype="GO", columns=c("SYMBOL", "ENTREZID"))

# find gene ontology tags for related to ZNF658, which has the specified ENTREZID
select(org.Hs.eg.db, keys= "26149", keytype="ENTREZID", columns="GO")

# save GO tags to a character vector
select(org.Hs.eg.db, keys= "26149", keytype="ENTREZID", columns="GO")$"GO"
# myk = .Last.value

# FOR KNIT purpose
# find GO tags related to ZNF658 (ENTREZID 26149)
go_tags <- select(org.Hs.eg.db,
                  keys = "26149",
                  keytype = "ENTREZID",
                  columns = "GO")
myk <- unique(na.omit(as.character(go_tags$GO)))

# identify biological processes ZNF658 is involved in
select(GO.db, keys=myk, columns="TERM")
```
### Assessment
```{r}
genes_17q21 <- select(org.Hs.eg.db,
                       keys = "17q21.1",
                       columns = "SYMBOL",
                       keytype = "MAP")

length(genes_17q21$SYMBOL)

m17 = select(org.Hs.eg.db, keys="17q21.1", keytype="MAP", columns=c("SYMBOL", "GO"))
sort(table(m17$GO), decreasing=TRUE)[1:10]

m17[m17$SYMBOL == "ORMDL3" & m17$EVIDENCE == "TAS", ]
```

### Assessment
```{r}
library(Homo.sapiens)
g = genes(Homo.sapiens)
library(ERBS)
data(HepG2)

# resize(g,1) is a strand-respecting reduction of the gene interval to a single base representing the transcription start site
kp = g[resize(g,1) %over% HepG2]

nn = names(kp)
m = select(Homo.sapiens, keys=nn, keytype="ENTREZID",
           columns=c("SYMBOL", "GENENAME", "TERM", "GO"))
library(DT)
datatable(m)
```
## Using Kyoto Encyclopedia of Genes and Genomes (KEGG)
```{r}
# load KEGGREST package and inspect organism-specific gene pathways
library(KEGGREST)
brca2K = keggGet("hsa:675")    # reference to a specific gene
names(brca2K[[1]])
brpat = keggGet("path:hsa05212")    # info on a pathway
brpat[[1]]$GENE[seq(1,132,2)] # entrez gene ids for pathway

# inspect some entrez ids
select(org.Hs.eg.db, keys="5888", keytype="ENTREZID", columns ="SYMBOL")
select(org.Hs.eg.db, keys="675", keytype="ENTREZID", columns ="SYMBOL")

# diagram showing structure of network
library(png)
library(grid)
brpng = keggGet("hsa05212", "image")
grid.newpage()
grid.raster(brpng)

# # working one
# url <- "https://rest.kegg.jp/get/hsa05212/image"
# tmp <- tempfile(fileext = ".png")
# 
# download.file(url, tmp, mode = "wb", quiet = TRUE)
# 
# img <- png::readPNG(tmp)
# grid::grid.newpage()
# grid::grid.raster(img)
# 
# # working one 2
# brpng <- keggGet("hsa05212", "image")
# 
# img <- if (is.list(brpng)) brpng[[1]] else brpng   # handle both behaviors
# 
# dim(img)        # should be 3 numbers, e.g. 925 1265 4
# range(img)      # should NOT be 0 0
# 
# grid.newpage()
# grid.rect(gp = gpar(fill = "white", col = NA))
# grid.raster(img)
```

### Assessment
```{r}
hsa_3845 <- unlist(keggGet("hsa:3845"))
hsa_3845$NAME
length(keggGet("hsa:3845")[[1]]$PATHWAY)

library(png)
oo = keggGet("hsa00790", "image")
writePNG(oo, "im.png")
grid.newpage()
grid.raster(oo)

```
## EMBL's ontology lookup service
```{r}
library(rols)
oo = Ontologies()
oo
oo[[1]]

glis = OlsSearch("glioblastoma")
glis
res = olsSearch(glis)
resdf = as(res, "data.frame") # get content
resdf[1:4,1:4]
resdf[1,5]  # full description for one instance
```
### Assessment
```{r}
library(rols)
diab = OlsSearch("diabetes")
olsRows(allRows(diab))

fulld = olsSearch(allRows(diab))
adf = as(fulld, "data.frame")
sort(table(adf$ontology_name), decreasing=TRUE)[1:10]

library(DT)
datatable(adf)


```

# Verified assessment
```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(org.Hs.eg.db)
library(ensembldb)
library(EnsDb.Hsapiens.v75)
library(GO.db)
library(AnnotationHub)
library(rtracklayer)
library(KEGGREST)

library(BSgenome.Hsapiens.UCSC.hg19)
Hsapiens

sort(seqlengths(Hsapiens),decreasing=TRUE)

nchar(Hsapiens$chrX)/nchar(Hsapiens$chrY)


# The TP53 gene codes for the p53 protein, a key regulator of cell cycle and stress response that is the most commonly mutated gene in cancer. Investigate TP53 in the systems biology databases.

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene    # abbreviate

keytypes(org.Hs.eg.db)
class(org.Hs.eg.db)
org.Hs.eg.db

TP53 <- select(org.Hs.eg.db, keys="TP53", keytype="SYMBOL", columns ="ENTREZID") # TP53 ENTREZ ID number
gene_coords <- genes(txdb)[TP53$ENTREZID]
gene_coords # TP53 gene location
width(gene_coords)

# answers
tp53loc <- genes(txdb, filter=list(gene_id="7157"))
seqnames(tp53loc)
genebp = width(tp53loc)
# ------

# TP53 transcripts and exons in txdb
txs.7157 <- transcriptsBy(txdb, by="gene")$"7157"
length(txs.7157) # the number of transcripts

exon <- exons(txdb)  # GRanges of exons
exon.7157 <- subsetByOverlaps(exon, gene_coords)
length(exon.7157) # the number of exons
sum(width(reduce(exon.7157))) # the number of base pairs in the exon of TP53
sum(width(reduce(exon.7157)))/width(gene_coords) # proportion of base pairs in TP53 contained in exons

# answers
length(transcripts(txdb, filter=list(gene_id="7157")))
length(exons(txdb, filter=list(gene_id="7157")))
exonbp = sum(width(reduce(exons(txdb, filter=list(gene_id="7157")))))
exonbp
exonbp/genebp    # genebp from previous question
# -------

# TP53 transcript information from ensembldb
library(ensembldb)
library(EnsDb.Hsapiens.v75)
edb <- EnsDb.Hsapiens.v75    # abbreviate

gnf <- GeneNameFilter("TP53")
txs <- transcripts(
  edb,
  filter  = gnf,
  columns = c("tx_id", "tx_biotype", "tx_cds_seq_start", "tx_cds_seq_end")
) # TP53 transcripts in the EnsDb.Hsapiens.v75 database

txs[1]
txs$tx_cds_seq_start[1] # start of the coding sequence for the first transcript in the list

i <- which(txs$tx_biotype == "processed_transcript")
width(txs[i]) # length of specific transcript

# answers
Tx = transcripts(edb, filter = GeneNameFilter("TP53"))
Tx
Tx$tx_cds_seq_start[1]
ind = Tx$tx_biotype == "processed_transcript"
width(Tx[ind])
# ------

# GO tags associated with TP53
keytypes(org.Hs.eg.db)
tp53_go <- select(org.Hs.eg.db,
                  keys = "TP53",
                  columns = "GO",
                  keytype = "SYMBOL")
tp53_go # gene ontology tags associated with TP53
length(unique(tp53_go$GO)) # the number of unique GO tags for TP53

# Supporting evidence for TP53 annotations
# EVIDENCE column: what kind of experimental evidence supports the involvement of the gene in a certain process
sum(tp53_go$EVIDENCE == "IMP") # the number of TP53 processes inferred from a mutant phenotype ("IMP")

tbl <- sort(table(tp53_go$GO), decreasing = TRUE)
head(tbl) # GO tag with the most entries of supporting evidence

# Using GO.db to to find the biological role of TP53
keytypes(GO.db)
select(GO.db, keys = "GO:0051726", keytype = "GOID", columns = "TERM") # biological process of TP53 with the most types of supporting evidence

# TP53 location liftOver to hg38 genome build
library(AnnotationHub)
ah <- AnnotationHub()

ah_human <- subset(ah, species == "Homo sapiens")
query(ah, "TP53")

# chain file that specifies how features match across genome builds
ah.chain <- subset(ah, rdataclass == "ChainFile" & species == "Homo sapiens")
query(ah.chain, c("hg19", "hg38"))
ch <- ah[["AH14150"]]

# perform the liftOver
TP53_19L <- liftOver(gene_coords[1], ch)
TP53_19L

# answers
hg19_to_hg38 = ah[["AH14150"]]
library(rtracklayer)
liftOver(gene_coords, hg19_to_hg38)
# -----

# p53 signaling pathway from KEGG
# load KEGGREST package and inspect organism-specific gene pathways
library(KEGGREST)
p53 = keggGet("hsa04115")    # reference to a specific gene
names(p53[[1]])
length(p53[[1]]$GENE)/2

p53pat = p53[[1]]$GENE # gene symbol + description
type(p53pat)
sum(grepl("\\bcyclin\\b", p53pat))
# OR
length(grep("cyclin", p53pat))

p53pat[grep("cyclin", p53pat)]    # inspect gene names containing cyclin
```








