---
title: 'Section 2.4: Whole Genome Analysis'
author: "Haeun Lee"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Whole Genome Analysis
## Measuring Methylation from Sequencing
## Measuring Methylation from Sequencing Assessment
### Measuring Methylation from Sequencing Q1
Reduced Representation Bisulfite Sequencing or RRBS is an experimental technique widely used to manipulate the regions of the genome we measure. An enzyme is used to cut DNA at CCGG and the general idea is to filter out small or large molecules once DNA is cut. We can use Bioconductor tools to predict the size of these regions.

Load the genome package and create an object with the sequence for chr22:

```{r}
library("BSgenome.Hsapiens.UCSC.hg19")
chr22 <- Hsapiens[["chr22"]]
```

Now use the matchPattern() function to find all the locations in which CCGG occurs on chr22.

How many CCGG do we find on chr22?

```{r}
res = matchPattern("CCGG",chr22)
```

### Measuring Methylation from Sequencing Q2
Plot a histogram of the DNA fragment sizes after we cut at CCGG positions. How would you describe this distribution?

```{r}
size=diff(start(res))
hist(size)
hist(log10(size))
```

### Measuring Methylation from Sequencing Q3

A typical size to filter are DNA regions between 40 and 220 basepairs.

What proportion of the fragments created for chr22 are between 40 and 220 basepairs?

```{r}
mean(size >=40 & size <=220)
```

### Measuring Methylation from Sequencing Q4

If we try to sequence all of chromosome 22, we would need to sequence 51,304,566 bases. If instead we only sequence fragments of sizes between 40 and 220 basepairs, how many bases would we need to sequence?

```{r}
frag.index = which(size >=40 & size <=220)
sum(size[frag.index])
```

### Measuring Methylation from Sequencing Q5

Whole-genome bisulfite sequencing (WGBS) is another sequencing-based protocol for measuring DNA methylation in samples. In the remaining problems, we will take a look at WGBS data from a set of paired tumor and normal colon samples. You can download the data from GitHub repo (~40 Mb). More information about these data are available on GEO GSE46644. Here we use the .cov files from Bismark as input to the bsseq R/Bioconductor package to create a bsseq object. The data come from this paper (https://pubmed.ncbi.nlm.nih.gov/23925113/).

Step-by-step instruction on how to covert the raw FASTQ files to the files that we will be work on are available here (https://github.com/genomicsclass/colonCancerWGBS/blob/master/scripts/createObject.Rmd).

Let's start by the reading in the target information.

```{r}
## path to the downloaded file collection
path <- '/Users/iha-eun/Desktop/HarvardX/02. Case Studies in Functional Genomics/Section 2_DNA methylation/colonCancerWGBS-master'

## read in sample metadata table
targets <- read.table(file.path(path,"targets.txt"), header = TRUE, sep = "	")
targets
```

Now you will need the bsseq package to read in the sequencing data.

We load the methylation calls from our alignments into R. Once the data are loaded into R, we can use this package for further downstream analyses such as finding differentially methylated regions between our paired tumor and normal samples. This package assumes the following data have been extracted from the alignments.

1. genomic positions (chromosome and location) for methylation sites
2. M (Methylation) values, the number of reads supporting methylation covering each site
3. Cov (Coverage) values, the total number of reads covering each site

For illustrative purposes, we only consider the methylation sites on chromosome 22 from the .cov file.

Here, read in the six files:

```{r}
# BiocManager::install("bsseq")
library("bsseq")

## turn metadata into DataFrame w/ sample names as rownames
targets <- DataFrame(targets, row.names = as.character(targets$Run))

## specify path to files in same order as targets table
covfiles <- file.path(path, paste0(rownames(targets), ".chr22.cov"))

## read coverage files
colonCancerWGBS <- read.bismark(files = covfiles, rmZeroCov = TRUE,
                                colData = targets)
```

Take a look at the bsseq object and the phenotypic information about for sample:

```{r}
colonCancerWGBS

# phenotypic information
pData(colonCancerWGBS)

# granges object
granges(colonCancerWGBS)
```

Now, extract the coverage and the number of reads with evidence for methylation:

```{r}
cov <- getCoverage(colonCancerWGBS,type = "Cov") # coverage
m <- getCoverage(colonCancerWGBS,type = "M") # the number of reads with evidence for methylation
```

What proportion of the reported CpGs have some coverage in all sample?

```{r}
mean(apply(cov,1,min)>0) # minimum higher than 0, row by row
```

### Measuring Methylation from Sequencing Q6

Compute the total coverage (across all samples) for each CpG site. Plot it against location.

How would you describe coverage?

```{r}
total.cov = apply(cov,1,sum) # sum coverage of all samples
plot(start(granges(colonCancerWGBS)),total.cov)
abline(h=200,col = "red")
abline(h=300,col = "blue")

# OR

tot = rowSums(cov)
##there are some very large values
hist(tot)
loc= start(colonCancerWGBS)
##plot by pieces
for(i in 1:11){
  index=1:100000+100000*i ##very ad-hoc
  plot(loc[index],tot[index],cex=.5,ylim=c(0,300))
}
```

### Measuring Methylation from Sequencing Q7

Assuming cov>0, which of the following pairs gives us:
1) an estimate of methylation status at each CpG and
2) a quantity proportional to the standard error of this estimate?

m/cov and 1/sqrt(cov)

```{r}
gr = GRanges(seqnames="22",ranges=IRanges(start=43793678,end= 45022550))
index=granges(colonCancerWGBS)%over%gr
library(rafalib)
i=1
index2=which(index & cov[,i]>=5 & cov[,i]<=50)
x=start(colonCancerWGBS)[index2]
y=m[index2,i]/cov[index2,i]
w=sqrt(cov[index2,i])/7
plot(x,y,cex=w)
```



